Von Neumann-Morgenstern and Savage Theorems for Causal Decision Making

### 2020-08-29 Theory_and_decision/Draft/Theory_and_decision_draft.tex
subseccion sobre teoria de decision antes de los teoremas

Reinforcement Learning (RL) \citep{sutton1998reinforcement} is the standard framework for automated learning by interaction, in which an agent, located at some time $t$ and some state $s$ has to make a choice in order to maximize her long-run discounted accumulated reward. Optimal policies, which are prescriptions of which action must be taken at some state $s$, satisfy the Bellman Equations \citep{Puterman:1994:MDP:528623} and are the standard solution to the problem. Such solution is equivalent to maximizing \textit{expected utility} \citep{webb2007game}, which is the decision criterion prescribed by the most commonly used decision-making theories: von Neumann-Morgenstern's and Savage's. Such theories rely on \textit{associative} information, which is based on patterns found in data, rather than stronger causal relations.

@@

\subsection{Aim}
Our goal is to establish the foundations for decision making algorithms that rely on the existence of causal relations even though such relations are not previously known or fully observable to a decision maker such as the algorithms found, for example, in \citep{bareinboim2015bandits,lattimoreNIPS2016,sen2017identifying}. We provide here an implementable decision making criterion which we build over known results from von Neumman and Morgenstern, Savage and Pearl. With our result we expect to highlight the importance and applicability of causal relations in decision making and the wide variety of possible applications.

@@

Other Decision Making theories exist, such as Kahneman and Tversky's Prospect Theory \citep{kahneman1979prospect}, Gilboa's Case-Based Decision Theory \citep{gilboa1995case}, among others that are out of the scope of this work. In Section \ref{joyce_theory} we make reference to another theory, Joyce's Causal Decision Making \citep{joyce1999foundations}.
%We will later talk about another theory: Joyce's Causal Decision Making \citep{joyce1999foundations}. 
For further details on classical (non-causal) decision making, see \cite{bernardo2000bayesian} and \cite{gilboa2009decision}.

\subsection{Savage's Theorem}
\label{savage_theorem}
If a rational decision maker does not know the probabilities of obtaining certain outcomes and does not have a precise, objective quantification of her preferences (utility function), then it is Savage's Theorem \citep{savage1954the}, which gives a formal choosing criterion. Savage's result extends von Neumann-Morgenstern Theorem since it considers the case in which a rational decision maker does not know neither her utility function nor the probabilities to be used in order to obtain the expected values required for making choices according to the von Neumann-Morgenstern Theorem. The notation and setting we use here is the one used by \cite{bernardo2000bayesian}. Consider a set $\Omega$. Consider a countable set $\mathcal{A}$ of available actions; for each action $a_i \in \mathcal{A}$, a partition $E_i$ of $\Omega$ and a set of consequences $C_i$. Let $\mathcal{E}$ the union of every $E_i$, we assume $\mathcal{E}$ to be a $\sigma$-algebra of events, and $\mathcal{C}$ the union of all $C_i$.

\begin{definition}
An uncertain environment is the tuple $(\Omega, \mathcal{A},\mathcal{C},\mathcal{E})$. Where $\mathcal{A}$ is a non-empty set of available actions, $\mathcal{C}$ a set of consequences and $\mathcal{E}$ an algebra of events over $\Omega$. 
\end{definition}

@@

\cite{bernardo2000bayesian} derive the existence of a subjective probability measure from a set of \textit{rationality axioms}, where a decision maker has some mechanism of quantifying uncertainty in terms of real numbers within the $[0,1]$ interval, and then the classical Kolmogorov axioms are derived from such construction, and therefore all of the known machinery of Probability Theory. We now state Savage's Theorem. Details can be found in \citep{kreps1988choice,bernardo2000bayesian,gilboa2009decision}.

>>>>

Reinforcement Learning (RL) \citep{sutton1998reinforcement} is the standard framework for automated learning by interaction, in which an agent, located at some time $t$ and some state $s$ has to make a choice in order to maximize her long-run discounted accumulated reward. Optimal policies, which are prescriptions of which action must be taken at some state $s$, satisfy the Bellman Equations \citep{Puterman:1994:MDP:528623} and are the standard solution to the problem. Such solution is equivalent to maximizing \textit{expected utility} \citep{webb2007game}, which is the decision criterion prescribed by the most commonly used decision-making theories: von Neumann-Morgenstern's and Savage's. Such theories rely on \textit{associative} information, which is based on patterns found in data, rather than stronger causal relations. In fact, it can be argued that RL is not a causal problem in itself \cite{gonzalezsoto2019reinforcement}.

@@

\subsection{Aim}
Our goal is to establish the foundations for decision making algorithms that rely on the existence of causal relations even though such relations are not previously known or fully observable to a decision maker such as the algorithms found, for example, in \citep{bareinboim2015bandits,lattimoreNIPS2016,sen2017identifying, gonzalez2018playing}. We provide here an easily implementable decision making criterion which we build over known results from von Neumman and Morgenstern, Savage and Pearl. With our result we expect to highlight the importance and applicability of causal relations in decision making and the wide variety of possible applications.

@@

Other Decision Making theories exist, such as Kahneman and Tversky's Prospect Theory \citep{kahneman1979prospect}, Gilboa's Case-Based Decision Theory \citep{gilboa1995case}, among others that are out of the scope of this work. In Section \ref{joyce_theory} we make reference to another theory, Joyce's Causal Decision Making \citep{joyce1999foundations}. For further details on classical (non-causal) decision making, see \cite{bernardo2000bayesian} and \cite{gilboa2009decision}.
%We will later talk about another theory: Joyce's Causal Decision Making \citep{joyce1999foundations}.

\subsection{Some Theory}
\begin{definition}
An uncertain environment is the tuple $(\Omega, \mathcal{A},\mathcal{C},\mathcal{E})$. Where $\mathcal{A}$ is a non-empty set of available actions, $\mathcal{C}$ a set of consequences and $\mathcal{E}$ an algebra of events over $\Omega$. 
\end{definition}

@@

\subsection{Savage's Theorem}
\label{savage_theorem}
If a rational decision maker does not know the probabilities of obtaining certain outcomes and does not have a precise, objective quantification of her preferences (utility function), then it is Savage's Theorem \citep{savage1954the}, which gives a formal choosing criterion. Savage's result extends von Neumann-Morgenstern Theorem since it considers the case in which a rational decision maker does not know neither her utility function nor the probabilities to be used in order to obtain the expected values required for making choices according to the von Neumann-Morgenstern Theorem. The notation and setting we use here is the one used by \cite{bernardo2000bayesian}. Consider a set $\Omega$. Consider a countable set $\mathcal{A}$ of available actions; for each action $a_i \in \mathcal{A}$, a partition $E_i$ of $\Omega$ and a set of consequences $C_i$. Let $\mathcal{E}$ the union of every $E_i$, we assume $\mathcal{E}$ to be a $\sigma$-algebra of events, and $\mathcal{C}$ the union of all $C_i$. \cite{bernardo2000bayesian} derives the existence of a subjective probability measure from a set of \textit{rationality axioms}, where a decision maker has some mechanism of quantifying uncertainty in terms of real numbers within the $[0,1]$ interval, and then the classical Kolmogorov axioms are derived from such construction, and therefore all of the known machinery of Probability Theory. We now state Savage's Theorem. Details can be found in \citep{kreps1988choice,bernardo2000bayesian,gilboa2009decision}.
